Gradual Soundness: Lessons from Static Python

### 2021-10-22 ecoop-2022/paper.tex
ecoop: temporary note about lack of Py Int optimizations
ecoop: sec2 intro edits

\SP{} is a type checker and bytecode compiler.
It validates programs written in ordinary Python syntax
and generates type-specific bytecode for the Cinder JIT compiler.
Cinder is a fork of CPython 3.8 that introduces a method-based
JIT and several bytecode instructions.

Running \SP{} code is no different than standard Python.
The main Cinder executable generates bytecode and performs
optimizations on the fly, same as CPython.

Almost all Python programs can run under \SP{} with no changes.
The few exceptions are due to restrictions that \SP{} puts
on classes to promote fast-running bytecode.

\subsection{Example Program}

@@

\subsection{Language Pipeline}

\begin{enumerate}
  \item Pyre as an optional, but recommended pre-check. (Not necessary for the model.)
  \item SP type checker
  \item Custom bytecode, Cinder JIT
\end{enumerate}

@@

\subsection{Virtual Method Tables}

\SP{} adds vtables to Python.
[FILL]

@@

  \item \textbf{Dynamic Checked Types}.
    A \code{CheckedDict} cannot be instantiated with \code{Dynamic}
    keys or values.
    [FILL] no desire, and makes runtime checks much simpler.

@@

>>>>

% {Language Pipeline}
% \begin{enumerate}
%   \item Pyre as an optional, but recommended pre-check. (Not necessary for the model.)
%   \item SP type checker
%   \item Custom bytecode, Cinder JIT
% \end{enumerate}

\SP{} is a type checker and bytecode compiler.
It validates programs written in ordinary Python syntax
and generates type-specific bytecode for the Cinder runtime system.
Cinder is an extension of CPython 3.8 that adds a method-based
JIT compiler, virtual method tables, and several bytecode instructions.
In sum, the projects consists of both a new type system (\SP{})
and a new language (Cinder = Python extended to leverage types).

Running \SP{} code is no different than standard Python,
though it takes place on a different platform.
The main Cinder executable generates bytecode and performs
optimizations on the fly, same as CPython.
This is one of the main advantages of \SP{} over mypyc,
which implements type-directed optimizations for CPython
by compiling source code to C extension modules.

Almost all Python programs can run under \SP{} with no changes.
The few exceptions are due to restrictions that \SP{} puts
on classes to promote fast-running bytecode.
In general, switching to \SP{} requires minimal effort.
Any additional work that developers are willing to invest
in refining their types is rewarded by performance gains.

@@

@@

\subsection{Runtime Changes}

\begin{itemize}
  \item
    \textbf{VTables}
    Cinder adds virtual method tables to all classes.

  \item
    \textbf{Bytecode Optimization}
    (could use a full section)

  %\item
  %  \textbf{Shadow Frames}
  %  (CinderDoc/static_python.rst)
  %  magic import that signals to the Static
  %  Python compiler to enable “shadow frame” mode in the Cinder JIT. This
  %  improves performance of function calls by avoiding the creation of full
  %  Python frame objects until they are definitely needed (e.g. if an
  %  exception is raised.) In the future this should become default.

@@

  \item \textbf{Dynamic Checked Types}.
    A \code{CheckedDict} cannot be instantiated with \code{Dynamic}
    keys or values.
    [FILL] no desire, and makes runtime checks much simpler.
    %% NB can't do ChkDict[ChkDict[_,_],_] either but that's because ChkDict is unhashable for now

@@

Optimize Python integer operations (in addition to the ctypes).
Hard because syntax like \code{a + b} may be the result of either \code{a.\_\_add\_\_(b)}
or \code{b.\_\_radd\_\_(a)} depending on runtime types and behavior.
